WSClean: an implementation of a fast, generic wide-field imager

### 2014-02-10 articles/2014/Article-WSClean/wsclean.tex
Work on article

\subsection{W-snapshots} \label{sec:snapshot-imaging}
\begin{figure*}
\begin{center}
\includegraphics[width=8cm]{img/vela-normal-projection}
\includegraphics[width=8cm]{img/vela-zenith-projection}
\caption{13-min MWA observation of supernova remnants Vela and Puppis A. Left: normal projection centred on Puppis A, right: phase-rotated to zenith to reduce $w$-values, recentred on Puppis A during imaging. Both Stokes-I images have been made with WSClean using the Cotton-Schwab clean algorithm. No beam correction was applied. TODO time difference and more discussion.}
\label{fig:vela-projection-example}
\end{center}
\end{figure*}
As discussed, all-sky snapshot imaging with zenith as phase centre is quite efficient. However, if one is not interested in imaging the whole sky, and the direction of interest is far from zenith, the computational overhead of making all-sky images is undesirable.

@@

A recentred image is in the projection of the zenith tangent plane, and will need an additional regridding step before multiple snapshots can be added together. A recentred image can be stored in a Flexible Image Transport System (FITS) file, which allows setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}. Common viewers such as Kvis and DS9 support such FITS files and display their coordinates correctly. An example of the difference in projection is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13~min MWA observation imaged with WSClean. Cleaning an image that is in a different projection yields slightly different results, but qualitativily the two images are clearly of equal accuracy.

@@

@@

Panel~\ref{fig:timing-channels} shows performance versus the output frequency resolution. The cost of imaging at multiple frequencies is the cost of making an images times the number of frequencies. However, each separate imaging contains a part of the total number of visibilities. As shown before, FFTs become the dominant cost for small visibility sets, and performance and number of frequency images therefore have a linear relation, which is clearly visible from the plot.

It is important to point out that all these measurement were performed by imaging a single polarization. Making full-polarization images with WSClean currently requires four runs. Imaging four Stokes parameters with CASA costs slightly less than four times the time of imaging a single polarization. With the settings of Table~\ref{tbl:default-parameters}, the difference is a factor of $3.5$, and imaging XX and YY together costs $1.9$ times imaging just XX.

@@

\begin{table*}
\caption{Predicted imaging times for various surveys, based on multi-frequency synthesis with 5 major iterations observing for 1 hour at ZA=20\degree.  Predictions are for $w$-projection in CASA, $w$-stacking with WSClean and $w$-stacked 12$\times$5-min snapshots with WSClean, with approximately equal accuracy.}
\begin{tabular}{lrrrrrrrrrr|rrr}
 Configuration&$\lambda$&FWHM  & Beams & Ant & Res & $\max |\Delta \mathbf{x}|$ & $\max \Delta z$ & int. t & BW & $\Delta \nu$ & $w$-proj & $w$-stack & $w$-snaps \\
          & (m)       &       &       &     &     & (km)                       & (m)             & (s)    & (MHz) &  (kHz) &  &  & \\
\hline
 GLEAM    &  2   & 24.7\degree& 1    & 128 & 2'   & 2.9 & 5 & 2 & 32 & 40     &14h 02m & 1h 52 m & 1h 23m \\
 EMU      & 0.2  & 1\degree   & 30   &  36 & 10''  & 6 &  0 &10 & 300 & 20    &19h 49m &14h 12 m & 14h 13m\\
 MSSS low &  5   & 9.8\degree & 5    &  20 & 100'' & 5 &    &10 & 16 & 16     & 2h 30m & 0h 11 m & 0h 11m\\
 MSSS high&  2   & 3.8\degree & 5    &  40 & 120'' & 5 &    &10 & 16 & 16     & 2h 31m & 0h 38 m & 0h 39m\\
 AARTFAAC &  5   & 45\degree  & 1    & 288 &  20'  & 0.3& 1 & 1 &  7 & 24     & & & \\
 VLSS     & 4.1  & 14\degree  & 1    &  27 &  80'' &11.1& 0 &10 & 1.56 & 12.2 & 0h 32m & 0h 12 m & 0h 02m\\
 SKA-P1 AA inner&3& 1.2\degree& 480  &  35 & 140'' &2.5& 2 &1.2 & 380 & 1     &       &         &      \\
 SKA-P1 AA mid&3 & 1.2\degree & 480  &  50 & 3.5'' &100& 50 &1.2 & 380 & 1     &       &         &      \\
\end{tabular}
\end{table*}

@@

\section{Application of WSClean to the MWA}
For technical reasons, the MWA correlator splits observations in data products of typically a few minutes. This simplifies $w$-snapshot imaging, which can be done by phase rotating each snapshot to zenith with a possibly additional phase-centre shift.

TODO: calculate when to use w-snapshots: discuss at what is the optimal time to split measurement sets into when using recentering technique.
TODO: calculate optimum splitting time?

\subsection{Beam correction}
Because the MWA consists of fixed, beamformed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix. Alternatively, Mueller matrices can be used as well, but in general these result in more computations \citep{revisiting-me-i}. When assuming all individual antennas of the tiles are working properly, all tiles have the same beam. The beam can then be corrected in image space for all tiles at once with
\begin{equation}\label{eq:beam-correction}
 I(l, m) = J(l,m)^{-1} I'(l, m) \left(J(l,m)^*\right)^{-1},
\end{equation}
where each term is a $2\times2$ complex matrices, with $J$ the voltage beam and $^*$ denoting the conjugate transpose. Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such. However, this is not correct, because $X$ is defined as being parallel to the celestial equator.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters. The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry. To make proper MWA beam correction possible, WSClean can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images. Each polarization requires a separate run of the $w$-stacking algorithm. Once these six images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}. This method avoids expensive beam-correcting kernels during gridding, but only works because all tiles have the same beam. To apply the same method on heterogeneous arrays such as LOFAR, each set of correlations with a different combination of station beams will have to be imaged separately, which will increase the cost of the algorithm. The AWImager uses an intermediate method, and corrects a common dipole factor in image space and the phased-array beam factor in $uv$-space, which allows the gridding kernel to be smaller compared to correcting both in $uv$-space \citep{awimager-2013}.

>>>>

\subsection{W-snapshots} \label{sec:snapshot-imaging}
\begin{figure*}
\begin{center}
\includegraphics[width=8cm]{img/vela-normal-projection}
\includegraphics[width=8cm]{img/vela-zenith-projection}
\caption{13-min MWA observation of supernova remnants Vela and Puppis A. Left: normal projection centred on Puppis A, right: phase-rotated to zenith to reduce $w$-values, recentred on Puppis A during imaging. Both Stokes-I images have been made with WSClean using the Cotton-Schwab clean algorithm. No beam correction was applied. Computing cost was 60 min for the normal projected image and 41 min for the recentred image.}
\label{fig:vela-projection-example}
\end{center}
\end{figure*}
As discussed, all-sky snapshot imaging with zenith as phase centre is quite efficient. However, if one is not interested in imaging the whole sky, and the direction of interest is far from zenith, the computational overhead of making all-sky images is undesirable.

@@

A recentred image is in the projection of the zenith tangent plane, and will need an additional regridding step before multiple snapshots can be added together. A recentred image can be stored in a Flexible Image Transport System (FITS) file, which allows setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}. Common viewers such as Kvis and DS9 support such FITS files and display their coordinates correctly. An example of the difference in projection is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13~min MWA observation imaged with WSClean. Cleaning an image that is in a different projection yields slightly different results, but qualitativily the two images are clearly of equal accuracy. The wall-clock time for imaging is 60 min and 41 min for respectively normal projection and the recentred image. Of the 41 min, 3 min is spent on phase rotating the visibilities.

@@

\subsection{Beam correction}
Because the MWA consists of fixed, beamformed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix. Alternatively, Mueller matrices can be used as well, but in general these result in more computations \citep{revisiting-me-i}. When assuming all individual antennas of the tiles are working properly, all tiles have the same beam. The beam can then be corrected in image space for all tiles at once with
\begin{equation}\label{eq:beam-correction}
 I(l, m) = J(l,m)^{-1} I'(l, m) \left(J(l,m)^*\right)^{-1},
\end{equation}
where each term is a $2\times2$ complex matrices, with $J$ the voltage beam and $^*$ denoting the conjugate transpose. Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such. However, this is not correct, because $X$ is defined as being parallel to the celestial equator.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters. The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry. To make proper MWA beam correction possible, WSClean can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images. Each polarization requires a separate run of the $w$-stacking algorithm. Once these six images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}. This method avoids expensive beam-correcting kernels during gridding, but only works because all tiles have the same beam. To apply the same method on heterogeneous arrays such as LOFAR, each set of correlations with a different combination of station beams will have to be imaged separately, which will increase the cost of the algorithm excessively unless an a-projection kernel is used. The AWImager uses an intermediate method, and corrects a common dipole factor in image space and the phased-array beam factor in $uv$-space, which allows the gridding kernel to be smaller compared to correcting both in $uv$-space \citep{awimager-2013}.

@@

Panel~\ref{fig:timing-channels} shows performance versus the spectral output resolution. The cost of imaging multiple frequencies is the cost of making an image from fewer visibilities times the number of frequencies. FFTs become the dominant cost for such small visibility sets, and the number of imaged frequencies affect performance linearly.

%It is important to point out that all these measurement were performed by imaging a single polarization. Making full-polarization images with WSClean currently requires four runs. Imaging four Stokes parameters with CASA costs slightly less than four times the time of imaging a single polarization. With the settings of Table~\ref{tbl:default-parameters}, the difference is a factor of $3.5$, and imaging XX and YY together costs $1.9$ times imaging just XX.

@@

\begin{table*}
\caption{Predicted computational cost for various configurations, based on multi-frequency synthesis with 5 major iterations observing for 1 hour at ZA=20\degree.  Predictions are for $w$-projection in CASA, $w$-stacking with WSClean and $w$-stacked 12$\times$5-min snapshots with WSClean, with approximately equal accuracy.}
\begin{tabular}{lrrrrrrrrrr|rrr}
 Configuration&$\lambda$&FWHM  & Beams & Ant & Res & $\max |\Delta \mathbf{x}|$ & $\max \Delta z$ & int. t & BW & $\Delta \nu$ & $w$-proj & $w$-stack & $w$-snaps \\
          & (m)       &       &       &     &     & (km)                       & (m)             & (s)    & (MHz) &  (kHz) &  &  & \\
\hline
 GLEAM    &  2   & 24.7\degree& 1    & 128 & 2'   & 2.9 & 5 & 2 & 32 & 40     &14h 02m & 1h 52m & 1h 23m \\
 EMU      & 0.2  & 1\degree   & 30   &  36 & 10''  & 6 &  0 &10 & 300 & 20    &19h 49m &14h 12m & 14h 13m\\
 MSSS low &  5   & 9.8\degree & 5    &  20 & 100'' & 5 &    &10 & 16 & 16     & 2h 30m & 0h 11m & 0h 11m\\
 MSSS high&  2   & 3.8\degree & 5    &  40 & 120'' & 5 &    &10 & 16 & 16     & 2h 31m & 0h 38m & 0h 39m\\
 AARTFAAC &  5   & 45\degree  & 1    & 288 &  20'  & 0.3& 1 & 1 &  7 & 24     & 10h 47m & 4h 58m & 4h 58m \\
 VLSS     & 4.1  & 14\degree  & 1    &  27 &  80'' &11.1& 0 &10 & 1.56 & 12.2 & 0h 32m & 0h 12m & 0h 02m\\
 SKA-P1 AA inner&3& 1.2\degree& 480  &  35 & 140'' &2.5& 2 &1.2 & 380 & 1     & 3218 d & 1865 d  & 1865 d \\
 SKA-P1 AA mid&3 & 1.2\degree & 480  &  50 & 3.5'' &100& 50 &1.2 & 380 & 1     & 6658 d & 3843 d & 3844 d \\
\end{tabular}
\end{table*}

@@
